---
title: "Single-Cell ATAC-seq Allele-Specific Pipeline"
subtitle: "From 10x Preprocessing to Post-EM Clustering"
author: "Your Name / Lab"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Environment & Dependencies
First, we establish package requirements and baseline configurations.

```{r}
# Bioconductor Dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

required_bioc <- c("Signac", "GenomeInfoDb", "EnsDb.Mmusculus.v79")
new_bioc <- required_bioc[!(required_bioc %in% installed.packages()[,"Package"])]
if(length(new_bioc)) BiocManager::install(new_bioc)

# CRAN Dependencies
required_cran <- c("Seurat", "hdf5r", "dplyr", "ggplot2", "uwot", "tidyr", "tibble")
new_cran <- required_cran[!(required_cran %in% installed.packages()[,"Package"])]
if(length(new_cran)) install.packages(new_cran)

# Load libraries
library(Signac)
library(Seurat)
library(GenomeInfoDb)
library(EnsDb.Mmusculus.v79)
library(hdf5r)
library(dplyr)
library(GenomicRanges)
library(ggplot2)
library(uwot)
library(tidyr)
library(tibble)

# Global configuration paths
PATH_BIALLELIC <- "scATAC_preB_all/"
PATH_GENOME1   <- "scATAC_preB_all_genome1_biPeaks/"
PATH_GENOME2   <- "scATAC_preB_all_genome2_biPeaks/"
PATH_BED_LIST  <- "async_master_list.GRCm38.noChr.bed"

# Fetch gene annotations for GRCm38/mm10
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)
#seqlevelsStyle(annotations) <- 'UCSC'
```


Custom Functions
We wrap repeated code into functions to optimize clarity and reproducibility.

```{r}
# Function to parse and load raw 10x datasets
import_atac <- function(count_path, meta_path, fragment_path) {
  counts <- Read10X_h5(filename = count_path)
  
  # Structural control: remove cells expressing exclusive Y chromosome signatures
  counts_Y <- counts[grep("Y", row.names(counts)), ]
  counts.keep <- counts[, apply(counts_Y, 2, function(x) all(x == 0))]
  
  meta <- read.csv(file = meta_path, header = TRUE, row.names = 1)
  
  chrom_assay <- CreateChromatinAssay(
    counts = counts,
    sep = c(":", "-"),
    genome = 'GRCm38',
    fragments = fragment_path,
    min.cells = 10,
    min.features = 10
  )
  
  data <- CreateSeuratObject(
    counts = chrom_assay,
    assay = "peaks",
    meta.data = meta
  )
  
  Annotation(data) <- annotations
  data <- NormalizeData(data, normalization.method = "LogNormalize", scale.factor = 10000)
  data <- NucleosomeSignal(object = data)
  data <- TSSEnrichment(object = data, fast = FALSE)
  
  data$blacklist_ratio <- data$blacklist_region_fragments / data$peak_region_fragments
  data$pct_reads_in_peaks <- data$peak_region_fragments / data$passed_filters * 100 
  
  return(data)
}

# Convert text-based regions back into GenomicRanges objects
convert_peaks_to_granges <- function(peaks) {
  peaks.df <- as.data.frame(do.call(rbind, lapply(peaks, function(x) unlist(strsplit(x, "[:-]")))))
  colnames(peaks.df) <- c("seqnames", "start", "end")
  makeGRangesFromDataFrame(peaks.df)
}

# Coordinate filtration across regional boundaries
filter_seurat_by_overlap <- function(seurat_obj, target_gr) {
  peaks <- row.names(seurat_obj[["peaks"]]@counts)
  peaks_gr <- convert_peaks_to_granges(peaks)
  overlap_indices <- IRanges::`%over%`(peaks_gr, target_gr)
  return(subset(seurat_obj, features = peaks[overlap_indices]))
}

# Track pristine cross-allelic unique hits
find_unique_overlaps <- function(gr1, gr2) {
  overlaps <- findOverlaps(gr1, gr2)
  overlaps <- overlaps[!duplicated(queryHits(overlaps)), ]
  overlaps <- overlaps[!duplicated(subjectHits(overlaps)), ]
  return(overlaps)
}

filter_seurat_by_hits <- function(seurat_obj, hits, type = "query") {
  indices <- if(type == "query") queryHits(hits) else subjectHits(hits)
  return(subset(seurat_obj, features = row.names(seurat_obj[["peaks"]]@counts)[indices]))
}

# Coordinate mapping and structural collapsing across the Master list
compute_sums_for_overlaps <- function(seurat_obj, gr, asrt.gr) {
  overlaps <- findOverlaps(gr, asrt.gr)
  result <- rowsum(as.matrix(seurat_obj[["peaks"]]@counts[queryHits(overlaps), ]), 
                   group = subjectHits(overlaps))
  
  sums_matrix <- as(result, "dgCMatrix")
  subset_asrt <- asrt.gr[unique(subjectHits(overlaps)), ]
  rownames(sums_matrix) <- paste0(seqnames(subset_asrt), "-", start(subset_asrt), "-", end(subset_asrt))
  
  chrom_assay <- CreateChromatinAssay(counts = sums_matrix, sep = c(":", "-"), genome = 'GRCm38')
  return(CreateSeuratObject(counts = chrom_assay, assay = "peaks"))
}

# Conjoin matching cells by rows across parallel matrices
merge_seurat_objects_by_rows <- function(seurat1, seurat2) {
  counts1 <- seurat1[["peaks"]]@counts
  counts2 <- seurat2[["peaks"]]@counts
  row.names(counts2) <- paste0("chr", row.names(counts2))
  
  if (!identical(colnames(counts1), colnames(counts2))) {
    stop("Cell barcodes do not match between genomic representations.")
  }
  
  combined_counts <- rbind(counts1, counts2)
  chrom_assay <- CreateChromatinAssay(counts = combined_counts, sep = c(":", "-"), genome = 'GRCm38')
  return(CreateSeuratObject(counts = chrom_assay, assay = "peaks"))
}
```

Phase 1: Preprocessing & Data Extraction
Here we ingest our primary biallelic baseline profile and compute strict dynamic distribution bounds to purge lower-tier elements.

```{r}
# Load baseline profiling controls
biallelic <- import_atac(
  paste0(PATH_BIALLELIC, "filtered_peak_bc_matrix.h5"),
  paste0(PATH_BIALLELIC, "singlecell.csv"),
  paste0(PATH_BIALLELIC, "fragments.tsv.gz")
)

# Statistical quantile dynamic evaluation
low_prf <- quantile(biallelic[["peak_region_fragments"]]$peak_region_fragments, probs = 0.02)
hig_prf <- quantile(biallelic[["peak_region_fragments"]]$peak_region_fragments, probs = 0.98)
low_prp <- quantile(biallelic[["pct_reads_in_peaks"]]$pct_reads_in_peaks, probs = 0.02)
hig_ns  <- quantile(biallelic[["nucleosome_signal"]]$nucleosome_signal, probs = 0.98)
low_ts  <- quantile(biallelic[["TSS.enrichment"]]$TSS.enrichment, probs = 0.02)

# Purge outlying cell traces
data_clean <- subset(
  x = biallelic,
  subset = peak_region_fragments > low_prf &
    peak_region_fragments < hig_prf &
    pct_reads_in_peaks > low_prp &
    nucleosome_signal < hig_ns &
    TSS.enrichment > low_ts
)

# Dereference background memory allocations
biallelic <- NULL
```


Allele Integration
Next, we map individual parental genomes (Genome1 and Genome2) across matching structural identities.

```{r}
allele1 <- import_atac(paste0(PATH_GENOME1, "filtered_peak_bc_matrix.h5"), 
                       paste0(PATH_GENOME1, "singlecell.csv"), 
                       paste0(PATH_GENOME1, "fragments.tsv.gz"))
allele2 <- import_atac(paste0(PATH_GENOME2, "filtered_peak_bc_matrix.h5"),
                       paste0(PATH_GENOME2, "singlecell.csv"), 
                       paste0(PATH_GENOME2, "fragments.tsv.gz"))

# Synchronize overlapping barcode matrices
cells.to.keep <- intersect(row.names(data_clean@meta.data), row.names(allele1@meta.data))
cells.to.keep <- intersect(cells.to.keep, row.names(allele2@meta.data))

allele1 <- subset(allele1, cells = cells.to.keep)
allele2 <- subset(allele2, cells = cells.to.keep)
```


Genomic Intersection Analysis
Filter elements against structural replication targets defined by our target BED annotations.

```{r}
asrt <- read.delim(PATH_BED_LIST, header = FALSE) %>%
  mutate(V1 = sub("^chr", "", V1)) %>%
  dplyr::filter(V1 != "X") %>%
  arrange(V1, V2, V3)
colnames(asrt) <- c("seqnames", "start", "end", "early")
asrt.gr <- makeGRangesFromDataFrame(asrt, keep.extra.columns = TRUE)

# Apply intersections
allele1 <- filter_seurat_by_overlap(allele1, asrt.gr)
allele2 <- filter_seurat_by_overlap(allele2, asrt.gr)

# Extract reciprocal unique representations
overlaps <- find_unique_overlaps(granges(allele1), granges(allele2))
allele1_asrt_overlap_allel2 <- filter_seurat_by_hits(allele1, overlaps, type = "query")
allele2_asrt_overlap_allel1 <- filter_seurat_by_hits(allele2, overlaps, type = "subject")

# Compute compressed master maps
c1_sums <- compute_sums_for_overlaps(allele1_asrt_overlap_allel2, granges(allele1_asrt_overlap_allel2), asrt.gr)
c2_sums <- compute_sums_for_overlaps(allele2_asrt_overlap_allel1, granges(allele2_asrt_overlap_allel1), asrt.gr)

# Construct final exported dataset matrices
combined <- merge_seurat_objects_by_rows(c1_sums, c2_sums)
c_merged <- combined[["peaks"]]@data
```

Exporting Targets
We write out structured inputs for classification and cache the workspace configuration state.

```{r}

tmp <- mergeByOverlaps(granges(c1_sums), asrt.gr)
tmp <- tmp[!duplicated(tmp[,1]), ]
early <- tmp$early

regions_numeric1 <- as.numeric(early == "CAST")
regions_numeric2 <- as.numeric(early != "CAST")
regions_numeric_merged <- c(regions_numeric1, regions_numeric2)

# Save deliverables to filesystem
write.csv(t(as.matrix(c_merged)), "c_merged_bi_preB.csv")
write.table(regions_numeric_merged, "regions_numeric_merged_bi_preB.csv", sep=",", row.names = FALSE, col.names = FALSE)

save.image("scATAC_preB_all_bi.RData")

```

